---
title: "Honest maps: classification, missingness and distortion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Honest maps: classification, missingness and distortion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 3.6, fig.align = "center", dpi = 96
)
library(countryatlas)
library(ggplot2)
snap <- world_snapshot$countries
has_maps <- requireNamespace("maps", quietly = TRUE)
has_sf <- requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("rnaturalearth", quietly = TRUE) &&
  requireNamespace("rnaturalearthdata", quietly = TRUE)
```

```{r setup-data, eval = has_maps}
mapdf <- attach_geometry(snap, geometry = "polygon")
```

A world choropleth makes four claims before it says anything about your data: that
the classes it drew are the natural ones, that grey means nothing rather than
zero, that a rate over eleven thousand people is worth as much attention as one
over a billion, and that the shapes on screen are the shapes on the ground.
All four are usually false. This vignette is the tour of what `countryatlas`
gives you for each.

## 1. The classification is doing the talking

Brewer & Pickle (2002) ran 56 subjects over nine series of mortality maps and
found **quantiles** among the most accurately read classifications, with natural
breaks (Jenks) below 70% as accurate. That is the reverse of the common GIS
default, and it matters because the choice is not cosmetic: it decides what the
reader concludes.

`classify_compare()` draws the same data under several methods at once.

```{r classify, eval = has_maps, fig.height = 3.2, fig.alt = "GDP per capita under quantile, Jenks, equal-interval and pretty breaks."}
cmp <- classify_compare(mapdf, gdp_per_capita, ncol = 2)
cmp
```

The picture is only half of it. The counts are attached to the plot:

```{r classify-report, eval = has_maps}
attr(cmp, "countryatlas_classification")
```

Equal-interval and pretty breaks put over 90% of countries into a single class,
because GDP per capita is strongly right-skewed and the top of the range is one
country. A map like that is technically correct and communicates nothing.
Quantiles put roughly 38 countries in each class.

You do not need the comparison to get the report — any *classified*
`world_map()` will produce it. (A continuous colourbar has no classes, so
asking there returns nothing and says why.)

```{r one-report, eval = has_maps}
p <- world_map(mapdf, gdp_per_capita, style = "quantile",
               classification_report = TRUE)
attr(p, "countryatlas_classification")
```

Jenks still earns its place: on a strongly clustered distribution, quantiles will
split a natural group across two colours where Jenks keeps it together. The point
is to look, not to take the default.

## 2. Grey is not a value

The default no-data grey reads as "low" to a lot of readers, which is precisely
the wrong inference. `na_style` gives you three alternatives, and `"hatched"`
(via the optional `ggpattern`) is the one that survives both colour-blindness and
a black-and-white printer.

```{r na-style, eval = has_maps, fig.alt = "World choropleth with missing countries drawn in diagonal hatching."}
world_map(mapdf, co2_per_capita, style = "quantile",
          na_style = "hatched", footnote = "auto")
```

`footnote = "auto"` writes the coverage line into the caption, so the map cannot
quietly overstate what it covers.

When the missingness *is* the story, map it directly:

```{r coverage, eval = has_maps, fig.alt = "Map of which countries report CO2 per capita."}
coverage_map(mapdf, co2_per_capita)
```

`audit_coverage()` is the same question as a table, and is the better tool when
you want to act on the answer rather than look at it.

## 3. Small denominators shout

Any per-capita or per-100k figure computed over a tiny population is mostly
noise, and on a choropleth it gets exactly as much ink as a figure computed over
a billion people. There are two answers. The cartogram distorts geometry until
area matches the denominator; **value-by-alpha** (Roth, Woodruff & Johnson 2010)
leaves the geometry alone and spends *opacity* instead.

```{r vba, eval = has_maps, fig.alt = "Value-by-alpha map: GDP per capita in colour, population as opacity, over a dark background."}
value_by_alpha_map(mapdf, gdp_per_capita, population)
```

Countries fade toward the background in proportion to how little population
stands behind their number. Compare with the cartogram answer to the same
problem, `cartogram_map()` / `dorling_map()`, in *Beyond the choropleth*: the
trade-off is that a cartogram makes the weighting unmissable but costs you the
recognisable world.

## 4. The projection is doing the talking too

Every flat world map distorts something. `projection_info()` says what each of
the thirteen preserves:

```{r proj-info}
projection_info()[, c("projection", "property", "equal_area", "conformal")]
```

For a choropleth the honest choice is **equal-area**, because the eye reads
coloured area as quantity — a projection that inflates Greenland makes
Greenland's value look more important than it is. Equal Earth is the package
default and the recommendation (Šavrič, Patterson & Jenny 2019).

```{r equal-area}
subset(projection_info(), equal_area)$projection
```

Tissot's indicatrix makes the cost visible. Each circle has the same radius on
the ground; whatever the projection does to them, it is doing to your data.

```{r tissot-merc, eval = has_sf, fig.height = 4, fig.alt = "Tissot indicatrices on Mercator: circles stay circular but grow enormously toward the poles."}
tissot_map("mercator")
```

```{r tissot-ee, eval = has_sf, fig.alt = "Tissot indicatrices on Equal Earth: ellipses shear but hold constant area."}
tissot_map("equal_earth")
```

Mercator keeps every circle round — it is conformal, so local shapes are right —
and grows them without limit toward the poles. Equal Earth keeps every circle's
*area* and shears the shapes instead. Neither is wrong; they are answers to
different questions, and only one of them belongs under a choropleth.

To see it on your own data, vary the CRS and hold everything else fixed:

```{r proj-compare, eval = has_sf, fig.height = 4.2, fig.alt = "One choropleth drawn under four projections."}
attach_geometry(snap, geometry = "sf") |>
  projection_compare(gdp_per_capita, style = "quantile", labeller = "property")
```

## 5. Say what the map is

Everything above is a decision, and a published map should carry its decisions.
`map_provenance()` reads them back off the plot.

```{r provenance, eval = has_maps, message = TRUE}
world_map(mapdf, gdp_per_capita, style = "quantile", n_bins = 5,
          na_style = "hatched", footnote = "auto") |>
  map_provenance()
```

Every field there was already known when the plot was built; the only new thing
is that you can read it. Paired with `footnote = "auto"` on the plot itself and
the classification report, that is most of a methods note.

Finally, `citation("countryatlas")` produces the package citation *and* the
sources it reconciles — `countrycode`, the World Bank, Natural Earth, and the
papers behind the methods used here. Citing the join layer without the data
would be the last dishonest thing a map could do.

## 6. Where the data comes from, and when

Two more ways a country map goes quietly wrong, both added in 3.0.0.

**The borders are not the borders.** A 1950 map drawn on 2024 boundaries is
simply a different world. `attach_geometry(year = )` and
[historical_geometry()] draw the real ones, from CShapes -- including the
colonies, without which most of Africa and Asia is absent:

```{r hist, eval = requireNamespace("cshapes", quietly = TRUE) && has_sf, fig.alt = "Choropleth drawn on 1950 borders including colonies and dependencies."}
attach_geometry(snap[, c("iso3c", "gdp_per_capita")], year = 1950) |>
  world_map(gdp_per_capita, style = "quantile",
            title = "1950 borders, 1950 world")
```

Note what this costs: ISO 3166 was published in 1974 and never covered
colonies, so historical geometry is keyed on Gleditsch-Ward codes and `iso3c`
is `NA` for every entity that never had one. `country_join(key = "gwn")` is the
join that works before 1970.

**Membership is a function of time.** A snapshot silently misstates any panel
that spans an accession:

```{r asof}
c(`2016` = in_group("United Kingdom", "EU", as_of = 2016),
  `2021` = in_group("United Kingdom", "EU", as_of = 2021))
```

## 7. Islands are not missing at random

`morans_i()`'s default weights are land-border contiguity, and an island has no
land border. On the bundled snapshot that silently removes a quarter of the
countries with data -- Japan, Australia, Madagascar, New Zealand, the
Philippines, Cuba, Sri Lanka, Iceland and every small island state. They are not
a random quarter.

Not every island goes, either: the United Kingdom keeps its land border with
Ireland, and Indonesia keeps its borders with Malaysia, Papua New Guinea and
Timor-Leste. Which is rather the point -- you cannot tell from the finished map
who dropped out of the statistic.

```{r weights, eval = has_sf}
rbind(
  contiguity = morans_i(snap, gdp_per_capita, n_perm = 0)[c("i", "n", "n_excluded")],
  knn = morans_i(snap, gdp_per_capita, n_perm = 0,
                 weights = country_weights("knn", k = 5))[c("i", "n", "n_excluded")]
)
```

Both numbers are defensible; only one of them is global. `country_weights()`
also takes `"distance"`, and `"custom"` -- which is how an adjacency that is not
geographic at all (trade volume, migration, shared language) goes through the
same API.

## References

Brewer, C. A. & Pickle, L. (2002). Evaluation of methods for classifying
epidemiological data on choropleth maps in series. *Annals of the Association of
American Geographers* 92(4), 662–681.

Roth, R. E., Woodruff, A. W. & Johnson, Z. F. (2010). Value-by-alpha maps: an
alternative technique to the cartogram. *The Cartographic Journal* 47(2),
130–140.

Šavrič, B., Patterson, T. & Jenny, B. (2019). The Equal Earth map projection.
*International Journal of Geographical Information Science* 33(3), 454–465.

