---
title: "Panels done right"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Panels done right}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = TRUE,
                      warning = TRUE)
library(countryatlas)
```

A cross-section has one row per country. A panel has one row per country and
year, and almost every country fact is a fact about a *date*: who belonged to
the EU, which income group a country was in, which region the World Bank put it
in. Treating those as fixed attributes of a country quietly paints today's world
over every year of the panel. This vignette walks through the verbs that keep
the dates attached. Everything here runs offline.

## A small panel

```{r}
pan <- expand.grid(iso3c = c("GBR", "FRA", "VNM", "PAK", "AFG"),
                   year = 2019:2026, stringsAsFactors = FALSE)
pan$gdp <- round(1000 * (1 + 0.03)^(pan$year - 2019) *
                   c(GBR = 42, FRA = 39, VNM = 3.5, PAK = 1.5, AFG = 0.5)[pan$iso3c])
head(pan)
```

## Joining two panels

`country_join()` reconciles the country keys on both sides. When both tables
carry a `year`, joining on the country alone would pair every year of one with
every year of the other, so it joins on the year too, and says so:

```{r}
pop <- expand.grid(iso3c = c("United Kingdom", "France", "Viet Nam"),
                   year = 2019:2026, stringsAsFactors = FALSE)
pop$population <- 1e6 * c(67, 68, 98)[match(pop$iso3c, unique(pop$iso3c))]
joined <- country_join(pan, pop, iso3c, iso3c, origin_x = "iso3c")
nrow(joined) == nrow(pan)
```

Pass `also_by = character()` to join on the country alone when one table really
is a cross-section to broadcast, and a named vector (`also_by = c(year = "yr")`)
when the year columns are named differently.

## Membership as of each row

`in_group()` takes one date per row, so each row asks about its own year. The
United Kingdom left the EU on 31 January 2020:

```{r}
uk <- pan[pan$iso3c == "GBR", ]
uk$eu <- in_group(uk$iso3c, "EU", origin = "iso3c", as_of = uk$year)
uk[, c("year", "eu")]
```

A bare year means 1 January of that year, so the UK is a member on 1 January
2020 and not on 1 January 2021. Suspensions are spells of their own: Syria is
not counted in the Arab League from 2011 to 2023.

```{r}
in_group(rep("SYR", 3), "ArabLeague", origin = "iso3c", as_of = c(2010, 2015, 2024))
```

## Classifications as they were

The World Bank classifies every economy once a year, on 1 July, from its gross
national income two years earlier. `classify_countries()` reads the dated table,
so each row gets the class in force at its date:

```{r}
classify_countries(pan[pan$iso3c == "VNM" & pan$year >= 2024, ], "income")
```

Viet Nam was lower middle income until 30 June 2026. The class *computed from* a
year's income is published two fiscal years later: Viet Nam's 2025 income put it
in the upper middle group, in force from 1 July 2026.

```{r}
classify_countries(data.frame(iso3c = "VNM", year = 2025), "income",
                   basis = "data_year")
```

## The region that moved

On 1 July 2025 the World Bank moved Afghanistan and Pakistan from South Asia
into its Middle East and North Africa region, renamed "Middle East, North Africa,
Afghanistan & Pakistan". A panel classified with today's list puts them there in
2019 too:

```{r}
reg <- classify_countries(pan[pan$iso3c %in% c("PAK", "AFG") &
                                pan$year %in% c(2025, 2026), ], "region")
reg[, c("iso3c", "year", "region")]
```

## Lags keyed on the year

`lag_by_country()`, `diff_by_country()` and `growth_rate()` look up the value a
year earlier, not the previous row. On a gapped panel the change across the gap
is `NA` rather than a two-year change under a one-year label:

```{r}
gappy <- pan[pan$iso3c == "FRA" & pan$year != 2022, c("iso3c", "year", "gdp")]
growth_rate(gappy, gdp)[, c("year", "gdp_growth")]
```

`by = "row"` keeps the old behaviour, for a panel that is irregular by design,
and warns about the gaps.

## Aggregates that know what they are missing

A regional average over the countries that happen to report is a different
number from the region's average. `aggregate_regions()` and `aggregate_groups()`
report how much of each group is behind each value, and leave a group out when
too little of it is (`min_coverage`, two thirds by default):

```{r}
aggregate_groups(world_snapshot$countries, gdp_per_capita,
                 groups = c("EU", "ASEAN", "SADC"), fun = "weighted_mean",
                 weight = population)
```

The coverage columns say which share of the group's members, and of its
population when a weight is given, the number rests on. The small panel above
has two of the EU's 28 members in 2019, so its EU figure is refused rather than
passed off as the EU's:

```{r}
aggregate_groups(pan, gdp, groups = "EU", as_of = 2019)
```
