---
title: "Modern maps with sf & projections"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modern maps with sf & projections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4, fig.align = "center", dpi = 96
)
library(countryatlas)
library(ggplot2)
has_sf <- requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("rnaturalearth", quietly = TRUE) &&
  requireNamespace("rnaturalearthdata", quietly = TRUE)
has_globe <- requireNamespace("maps", quietly = TRUE) &&
  requireNamespace("mapproj", quietly = TRUE)
```

The legacy `maps` polygons are an unprojected plate carrée: they badly distort
area and split Russia, Fiji and New Zealand across the antimeridian. The `sf`
backend fixes all of this — real projections, equal-area options, and an
antimeridian-safe pipeline. These features require the optional `sf` and
`rnaturalearth` packages.

```{r eval = FALSE}
install.packages(c("sf", "rnaturalearth", "rnaturalearthdata"))
```

## An equal-area, projected choropleth

```{r eval = FALSE}
world_data(2020, c(gdp = "NY.GDP.PCAP.KD"), geometry = "sf") |>
  world_map(gdp, style = "quantile", projection = "equal_earth",
            title = "GDP per capita (Equal Earth projection)")
```

`world_map()` auto-detects the `sf` backend and applies the projection through
`ggplot2::coord_sf()`. Available projections are `"equal_earth"` (the default —
equal-area and good-looking), `"robinson"`, `"mollweide"`, `"natural_earth"`,
`"plate_carree"`, `"mercator"`, `"winkel_tripel"`, `"eckert4"`, `"gall_peters"`,
`"orthographic"`, `"azimuthal_equal_area"`, `"north_polar"` and `"south_polar"`.

`projection_info()` says what each one preserves, so you do not have to
remember:

```{r}
projection_info()[, c("projection", "property", "equal_area")]
```

For a choropleth the equal-area ones are the honest choice, because the reader
reads coloured area as quantity. `projection_compare()` draws your own data
under several at once and `tissot_map()` shows the distortion directly — see
the *Honest maps* vignette.

## The world as a globe

`globe_map()` draws an orthographic globe centred on `lon`/`lat`. The default
`"sf"` backend gives the cleanest limb; the `"polygon"` backend below needs only
`maps` + `mapproj` (no `sf`):

```{r globe, eval = has_globe, fig.width = 5.5, fig.height = 5, fig.alt = "Orthographic globe centred on the Indian Ocean, countries coloured by continent."}
globe_map(world_snapshot$countries, continent, backend = "polygon",
          style = "categorical", lon = 10, lat = 20)
```

```{r eval = FALSE}
# With the sf backend (smoother limb, real great circles):
world_data(2020, geometry = "sf") |>
  globe_map(gdp_per_capita, lon = 10, lat = 30)
```

`spin_globe()` turns that into a rotating animation — one `globe_map()` frame per
central longitude, assembled into a looping GIF with `gifski` (or `magick`):

```{r eval = FALSE}
spin_globe(world_snapshot$countries, continent, backend = "polygon",
           style = "categorical", n_frames = 60)
```

## Just the canvas

`world_geometry()` returns projected, region-subset, antimeridian-safe geometry
without any data — country polygons, label-ready centroids, coastlines, a
graticule or an ocean rectangle:

```{r eval = has_sf, fig.alt = "Africa drawn on its own under an Equal Earth projection."}
africa <- world_geometry("countries", geometry = "sf", region = "Africa",
                         projection = "equal_earth")
ggplot(africa) +
  geom_sf(fill = "grey85", colour = "grey40", linewidth = 0.1) +
  theme_world_map()
```

```{r eval = !has_sf, echo = FALSE, results = "asis"}
cat("> The live `sf` map is not shown because `sf` is not installed in this build.\n")
```

## Recentring and the antimeridian

A Pacific-centred world is one argument away; the `sf` pipeline runs
`sf::st_break_antimeridian()` before projecting, so nothing streaks across the
frame:

```{r eval = FALSE}
world_geometry("countries", geometry = "sf", recenter = 150)
```

## Region subsetting

`region` accepts a continent, a group name (`"EU"`, `"OECD"`, …), a vector of
`iso3c` codes, or a bounding box `c(xmin, ymin, xmax, ymax)`. Pair it with a
`projection` suited to the subset (e.g. `"azimuthal_equal_area"` for a single
continent, the polar projections for the Arctic) so the crop stays area-honest.

A box is the one form that clips the shapes themselves, which is what you want
for a region that is neither a continent nor a group — the Mediterranean basin,
say. Use the `sf` backend for it: that clip is a real `sf::st_crop()`, whereas
the polygon backend can only drop the vertices outside the box and warns that a
country crossing the edge comes back with an approximate outline.

```{r eval = has_sf, fig.alt = "The Mediterranean basin cropped by bounding box under an Equal Earth projection."}
med <- world_geometry("countries", geometry = "sf",
                      region = c(-10, 30, 40, 48), projection = "equal_earth")
ggplot(med) +
  geom_sf(fill = "grey85", colour = "grey40", linewidth = 0.1) +
  theme_world_map()
```

## Simplifying for the web

High-resolution geometry can be thinned for fast plotting with
`simplify_geometry()` (which uses `rmapshaper` when available). `scale =
"medium"` is the 50m Natural Earth data; `"large"` is 10m and additionally
needs the `rnaturalearthhires` package, which is not on CRAN.

```{r eval = FALSE}
world_geometry(geometry = "sf", scale = "medium") |>
  simplify_geometry(keep = 0.1)
```
