---
title: "Accessible and citable maps"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 8, fig.height = 3.6, dpi = 96)
library(countryatlas)
p <- world_map(attach_geometry(world_snapshot$countries), gdp_per_capita)
```

A published map should say what it shows to a reader who cannot see it, be
readable with a colour-vision deficiency, and say where its numbers came from.

## Alt text

Every map verb sets its alt text from the final plot (Lundgard & Satyanarayan's
levels 1 and 2):

```{r}
map_alt_text(p)
```

knitr does not ask ggplot2 for it, so set it on the chunk from a plot built in
an earlier chunk, as this page does:

```{r, fig.alt = map_alt_text(p)}
p
```

## Colour vision

`check_palette()` simulates the three common deficiencies and reports the
smallest difference between adjacent classes:

```{r}
check_palette(p)
```

## Provenance and citation

```{r}
map_provenance(p)
cat(map_citation(p), sep = "\n\n")
```
