---
title: "The comparative-development toolkit"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, dpi = 96)
library(countryatlas)
snap <- world_snapshot$countries
```

## Inequality, concept by concept

Milanovic separates inequality between countries as units (concept 1) from
inequality between countries weighted by their populations (concept 2).
`inequality()` gives every standard measure under either:

```{r}
inequality(snap$gdp_per_capita)
inequality(snap$gdp_per_capita, weights = snap$population)
```

How much of it lies between continents? Theil's L decomposes without depending
on the order:

```{r}
theil(snap$gdp_per_capita, snap$population, groups = snap$continent, type = "L")
```

## Convergence and mobility

`beta_convergence()` and `sigma_convergence()` summarise a distribution in one
number each. `transition_matrix()` shows how countries move within it (Quah):

```{r}
set.seed(4)
pan <- expand.grid(iso3c = snap$iso3c[1:60], year = 2000:2010,
                   stringsAsFactors = FALSE)
pan$gdp <- exp(9 + stats::ave(stats::rnorm(nrow(pan), 0, 0.05), pan$iso3c,
                              FUN = cumsum) + rep(stats::rnorm(60), 11))
tm <- transition_matrix(pan, gdp, n_classes = 3, classes = "relative")
tm
attr(tm, "ergodic")
rank_mobility(pan, gdp, from = 2000, to = 2010)
```

## Money across countries and years

`deflate()` turns current prices into constant ones, and `to_ppp()` converts
exchange-rate dollars to purchasing-power parity; a cross-country panel over
time usually needs both. Both fetch their price series from the World Bank:

```{r, eval = FALSE}
d <- country_data(2010:2020, c(gdp = "NY.GDP.MKTP.CN"), panel = TRUE)
d |> deflate(gdp, base_year = 2015) |> to_ppp(gdp_real)
```
