---
title: "Disputed territories and worldviews"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, dpi = 96)
library(countryatlas)
```

Every world map takes a position on disputed territory, including the ones
that think they do not. The package takes none of its own: it records which
disputes your data touches, lets you state your convention, and draws a
published point of view when you choose one.

```{r}
check_dispute_coverage(world_snapshot$countries)
```

`dispute_policy()` states the convention, and the map verbs record it:

```{r}
old <- dispute_policy("neutral")
dispute_policy()
dispute_policy(old)
```

Natural Earth publishes its countries under 31 points of view at 1:10m. Set
one, by the viewing country's ISO code, and the `sf` backend draws it (the file
is downloaded once into the package cache):

```{r, eval = FALSE}
dispute_policy("de_jure", worldview = "IND")
world_map(attach_geometry(world_snapshot$countries, geometry = "sf"),
          gdp_per_capita)
dispute_policy(worldview = NA)                     # back to the default view
```
