---
title: "Rates and small numbers"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 8, fig.height = 3.6, dpi = 96)
library(countryatlas)
set.seed(3)
d <- world_snapshot$countries
d <- d[!is.na(d$population), c("iso3c", "population", "continent")]
d$deaths <- stats::rpois(nrow(d), d$population * 0.008 *
                           exp(stats::rnorm(nrow(d), 0, 0.2)))
```

A rate over eleven thousand people shouts as loudly as one over a billion. Four
verbs handle that: `rate_check()` flags it, `smooth_rates()` shrinks it,
`rate_funnel()` shows it, `value_by_alpha_map()` maps it.

```{r}
head(rate_check(d, deaths, population))
```

Empirical-Bayes smoothing pulls an unreliable rate toward the global rate,
or with `method = "local_eb"` toward its neighbourhood's:

```{r}
s <- smooth_rates(d, deaths, population, method = "local_eb")
head(s[order(s$population), c("iso3c", "deaths_rate", "deaths_smoothed",
                              "deaths_shrinkage")])
```

The funnel plot puts every country against the limits chance alone would give
a country its size, with Spiegelhalter's overdispersion adjustment:

```{r, fig.alt = "Funnel plot of a made-up death rate against population, with 95% and 99.8% control limits."}
rate_funnel(d, deaths, population, overdispersion = TRUE)
```
