---
title: "Spatial statistics on a world of islands"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 8, fig.height = 3.6, dpi = 96)
library(countryatlas)
snap <- world_snapshot$countries
```

A spatial statistic is only as good as its idea of who is next to whom. On a
world map, land borders leave Japan, Australia, New Zealand and every island
state with no neighbours at all, so a statistic on contiguity weights silently
drops a quarter of the countries with data. The package defaults to the five
nearest neighbours, so every country with a centroid takes part.

```{r}
knn <- morans_i(snap, gdp_per_capita, n_perm = 499)
border <- morans_i(snap, gdp_per_capita, n_perm = 499,
                   weights = country_weights("contiguity"))
rbind(knn = knn[, c("i", "n", "n_excluded", "p_value")],
      contiguity = border[, c("i", "n", "n_excluded", "p_value")])
```

## Local clusters, with the false discoveries controlled

A local Moran test runs once per country, so about one in twenty comes out
"significant" by chance at the 5% level. `local_morans()` and `lisa_map()`
control the false discovery rate by default (Caldas de Castro & Singer 2006),
on 9,999 conditional permutations so the smallest p-values can clear it.

```{r, fig.alt = "LISA cluster map of GDP per capita with false-discovery-rate control."}
lisa_map(attach_geometry(snap), gdp_per_capita, n_perm = 999)
```

Two variables at once: high GDP per capita among long-lived neighbours.

```{r}
b <- bivariate_lisa(snap, gdp_per_capita, life_expectancy, n_perm = 999)
table(b$cluster)
```

## Rates

Moran's I on raw rates mistakes the noise of small denominators for
clustering. `eb_morans_i()` standardises each rate by its own reliability
first (Assuncao & Reis 1999); it agrees with `spdep::EBImoran.mc()`.

```{r}
set.seed(2)
d <- snap[!is.na(snap$population), ]
d$births <- stats::rpois(nrow(d), d$population * 0.018)
eb_morans_i(d, births, population, n_perm = 499)
```

## Categories and dynamics

Do income groups cluster? Join counts compare the neighbouring pairs in the
same group with random labelling:

```{r}
join_counts(snap, income, n_perm = 499)
```

`spatial_markov()` asks whether a country's neighbours change its odds of
moving between classes (Rey 2001), with a homogeneity test; see
`?spatial_markov`.
